A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-10 text/manuscript.Rmd
updated fig4 annotation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(rstan)
library(foreach)
library(sqldf)
library(tidyverse)
library(ggthemes)
library(data.table)
library(biomaRt)
library(dplyr)
library(GenomicRanges)
library(SummarizedExperiment)
library(reactome.db)
library(plyr)
library(cluster)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(umap)
library(dbscan)
library(class)
library(TCGAbiolinks)
library(patchwork)
library(ggplotify)
library(ComplexHeatmap)
library(knitr)
library(kableExtra)
library(UpSetR)
```

@@

Other studies aim at providing a comprehensive view of mutations, gene
expression and genomic signatures, with the express goal of understanding the
common themes of all cancer independent of tissue of origin. Such a view of
cancer as a malady affecting common cellular processes has long been a goal of
the field and is perhaps best characterized in essays by @weinberg2000; @hallmarks_II (see also
Discussion). The first of these studies, (the
“Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types,
using integrative analysis with co-equal weighting of gene expression,
methylation, copy number and proteomics data
[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though a significant
proportion of tumors ($10\%$) could be reclassified independent of
tissue-of-origin. In a second study involving 33 cancer types and a much greater
number of tumors using a newer algorithm (iCluster)
[@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
could be further subdivided into organ specific groups, including
pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/
squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].

@@

```{r figure-4-data-generation, eval = FALSE}
tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/gene_data.Rda")

## Parameters ##
filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
n_path <- 10 ## looking at the top10 path

@@

toppath <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top10 mutated pathways

@@

```{r figure-4, eval = FALSE}
load("../data/class67_gene_2_path_heatmap.Rda") ## load data
load("../data/path_match_in_set.Rda")

@@

fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
  geom_tile(size = 0.6) +
  scale_fill_manual(values = c("grey", "firebrick3")) +
  #facet_grid(factor(path_combo, levels = most_genes$path_combo) ~ ., scales = "free", space = "free") +
  #scale_y_discrete(position = "left") +
  labs(tag = "A") +
  ylab("") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        strip.text.y = element_blank(),
        strip.background = element_blank(),
        legend.position = "none") 
fig4
ggsave("fig4_ggplot_heatmap.png", height = 100, width = 17, limitsize = FALSE)

@@

## Supplementary Information
```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data

@@

@@

![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]
<div style="text-align: right"> Supplementary Figure S3 Data </div>
```{r supplementary-figure-s3-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

@@

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(rstan)
library(foreach)
library(sqldf)
library(tidyverse)
library(ggthemes)
library(data.table)
library(biomaRt)
library(dplyr)
library(GenomicRanges)
library(SummarizedExperiment)
library(reactome.db)
library(plyr)
library(cluster)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(umap)
library(dbscan)
library(class)
library(TCGAbiolinks)
library(patchwork)
library(ggplotify)
library(ComplexHeatmap)
library(knitr)
library(kableExtra)
library(UpSetR)
library(cowplot)
```

@@

Other studies aim at providing a comprehensive view of mutations, gene
expression and genomic signatures, with the express goal of understanding the
common themes of all cancer independent of tissue of origin. Such a view of
cancer as a malady affecting common cellular processes has long been a goal of
the field and is perhaps best characterized in an essay by @weinberg2000 (see also
Discussion). The first of these studies, (the
“Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types,
using integrative analysis with co-equal weighting of gene expression,
methylation, copy number and proteomics data
[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though a significant
proportion of tumors ($10\%$) could be reclassified independent of
tissue-of-origin. In a second study involving 33 cancer types and a much greater
number of tumors using a newer algorithm (iCluster)
[@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
could be further subdivided into organ specific groups, including
pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/
squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].

@@

```{r figure-4-data-generation, eval = FALSE}
tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/moderate_enrich_prob.Rda")
load("../data/gene_data.Rda")

## Parameters ##
filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
n_path <- 10 ## looking at the top10 path
enrich_cutoff <- 0.30
################

@@

toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
  filter(clust %in% filtered_class) %>% ## filter the cluster
  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways

@@

```{r figure-4, message = FALSE}
load("../data/class67_gene_2_path_heatmap.Rda") ## load data
load("../data/path_match_in_set.Rda")

@@

fig4 <- ggplot(data = class67_gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
  geom_tile(size = 0.6) +
  scale_fill_manual(values = c("grey", "firebrick3")) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        strip.text.y = element_blank(),
        strip.background = element_blank(),
        legend.position = "none")

load("../data/path_match_in_set.Rda")

## right annotation data generation
proteasome_r_anno <- data.frame(gene = gene_order, 
                                psm_gene = 0)

proteasome_r_anno[grep(pattern = "PSM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1

r_anno_color <- c("0" = "white", "1" = "black")
r_anno <- ggplot(proteasome_r_anno) +
  geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
           stat = "identity", 
           width = 1) +
  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
  theme_void() 
legend <- plot_grid(get_legend(r_anno), ncol = 1)
r_anno <- r_anno + theme(legend.position = "none")

## left annotatoin data generation
l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap

## need to get top path names again
load("../data/moderate_enrich_prob.Rda")
enrich_cutoff <- 0.30
filtered_class <- c(6,7)
n_path <- 10
toppath <- moderate_enrich_prob %>% filter(signif <= 2500 & difference >= enrich_cutoff) %>% ## signif is 95% CI with a specified cutoff
  filter(clust %in% filtered_class) %>% ## filter the cluster
  arrange(desc(difference)) %>% ## arrange enrichment score in descending order
  .[, "path"] %>% as.character() %>% unique() %>% .[1:n_path] ## getting the top10 enriched pathways

for(p in toppath){
  message(p)
  l_anno <- l_anno %>% mutate(p = 0) ## create a new col named p
  names(l_anno)[names(l_anno) == 'p'] <- p ## rename col with pathway name
  ## for any path_combo that includes the pathway name change the 0 into a 1
  if(p == "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it has this pathway grep a different way
    l_anno[grep(pattern = "AUF1", x = l_anno$path_combo), p] <- 1
  }
  if(p != "AUF1 (hnRNP D0) binds and destabilizes mRNA") { ## if it is not AUF1 do it normally
  l_anno[grep(pattern = p, x = l_anno$path_combo), p] <- 1 
  }
}
l_anno_ggplot <- list() ## list for storing ggplots
path_legend <- list()
for(n in 1:length(toppath)){
  message(n)
  p <- toppath[n]
  message(p)
  spec_path <- l_anno[,c("gene_name", p)]
  names(spec_path)[names(spec_path) == p] <- "p"
  path_ggplot <- ggplot(spec_path) +
  geom_bar(mapping = aes(x = 1, y = gene_name, fill = as.character(p)), 
           stat = "identity", 
           width = 1) +
  scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = p) +
  theme_void() 
  path_legend[[n]] <- plot_grid(get_legend(path_ggplot), ncol = 1) ## extract legend from plot
  l_anno_ggplot[[n]] <- path_ggplot + theme(legend.position = "none") ## remove legend from plot
}

plot <- plot_grid(l_anno_ggplot[[1]], l_anno_ggplot[[2]], l_anno_ggplot[[3]], l_anno_ggplot[[4]], l_anno_ggplot[[5]], l_anno_ggplot[[6]], l_anno_ggplot[[7]], l_anno_ggplot[[8]], l_anno_ggplot[[9]], l_anno_ggplot[[10]], fig4, r_anno, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, 10), 15, 0.1))
plot

@@

## Supplementary Information

### Supplementary Figure S1

```{r supplementary-figure-s1, fig.width= 8.5, fig.height= 11}
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") ## load data

@@

### Supplementary Figure S2

@@

![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]

### Supplementary Figure S3

<div style="text-align: right"> Supplementary Figure S3 Data </div>
```{r supplementary-figure-s3-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

@@

### Supplementary Figure S4

@@

### Supplementary Table ST1
